Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-02-29 paper/paper.tex
Minor

More concretely, our paper makes the following contributions.
\begin{itemize}
\item In Section~\ref{sec:spec}, we introduce \lang{}, the dependent type
  theory of interest. A key design choice that impacts our proofs is the use
  of an untyped conversion rule, inspired by Pure Type
  Systems~\citep{barendregt1991introduction}, and specified through parallel
  reduction~\citep{takahashi-parallel-reduction,
    barendregt:lambda-calculi-with-types}.
\item In Section~\ref{sec:logreldep}, we formulate logical consistency for
  \lang{}, the property of interest, to motivate a logical relation. We define
  the relation first inductively and then later prove that it is a partial
  function. Based on this definition, we define semantic typing and prove the
  fundamental theorem, from which consistency follows as a corollary
  (Section~\ref{sec:logrelproof}). Thanks to the design of \lang{}, our proof
  showcases the special treatment required to model many of the most common
  features of dependent type theories, thus making our proof applicable to a
  broad range of type systems.
\item We strengthen our logical relation to prove the existence of
  $\beta$-normal forms (Section~\ref{sec:extension}) and $\beta\eta$-normal
  forms (Section~\ref{sec:betaeta}) for well-typed open terms. The
  modifications made to our initial logical relation are small and closely
  mirror the necessary extensions for a simply-typed language. We use this
  part to show that once we have established the base techniques, we can port
  ideas from proofs about simpler languages to the dependently typed setting.
\item We mechanize all our proofs using the Coq proof assistant, with 963
  lines of code for the consistency proof and a moderate increase to 1357
  lines of code for the normalization proof.  We discuss the specifics related
  to our choice of Coq as our metatheory, including our use of off-the-shelf
  semantic engineering infrastructure and automation tools, in
  Section~\ref{sec:logrelmech}.  Our proof scripts, with comments, are
  available to reviewers as supplementary materials.
\item We compare our work to existing proofs by logical relations and other
  proof techniques for proving consistency and normalization.  We provide an
  overview of this prior work (Section~\ref{sec:relatedwork}) and also give an
  in-depth explanation on how various design decisions affect the size of our
  proof and its extensibility to additional features
  (Section~\ref{sec:discuss}).
\end{itemize}

@@

The syntax of \lang can be found in Figure~\ref{fig:syntax}.  As a dependent
type theory, terms and types are collapsed into the same syntactic
category. The type $[[Set i]]$ represent a universe type, annotated by its
universe level, a natural number $i$.  Abstractions $[[ \ x . a ]]$ and
dependent function types $[[Pi x : A . B]]$ are binding forms for the variable
$x$ in the body of the function and codomain of the function type.
\footnote{In the exposition in this paper, binding forms are equal up to
  $\alpha$-conversion and we adopt the Barendregt Variable
  Convention~\cite{barendregt:lambda-calculus}, which lets us assume that
  bound variables are distinct.  In some places, we are informal about the
  treatment of variables and substitution; our mechanized proofs make these
  notions precise by using de Bruijn indices~\cite{debruijn1994automath}. }
We use the notation $[[A -> B]]$ when the output type $[[B]]$ is not dependent
on the input variable. For simplicity, we omit the type annotations in the abstraction
forms. We discuss how the inclusion of type annotations can affect our
our development in Section~\ref{sec:betaeta}, where we extend our
consistency result to the existence of $\beta\eta$-normal forms.
We include in \lang{} the intensional identity type
$[[a ~ b : A]]$ whose proofs can be eliminated by the J-eliminator
$[[J c a b p]]$, where $[[p]]$ is an equality proof between $[[a]]$ and
$[[b]]$, and $[[c]]$ is the term whose type is to be casted.  Finally, \lang{}
includes booleans, with standard syntax.  \scw{Example of what you can use
  identity types for? Or an example of a program that uses J?}

>>>>

More concretely, our paper makes the following contributions.
\begin{itemize}
\item In Section~\ref{sec:spec}, we introduce \lang{}, the dependent type
  theory of interest. A key design choice that impacts our proofs is the use
  of an untyped conversion rule, inspired by Pure Type
  Systems~\citep{barendregt1991introduction}, and specified through parallel
  reduction~\citep{takahashi-parallel-reduction,
    barendregt:lambda-calculi-with-types}.
\item In Section~\ref{sec:logreldep}, we formulate logical consistency for
  \lang{}, the property of interest, to motivate a logical relation. We define
  the relation first inductively and then later prove that it is a partial
  function. Based on this definition, we define semantic typing and prove the
  fundamental theorem, from which consistency follows as a corollary
  (Section~\ref{sec:logrelproof}). Thanks to the design of \lang{}, our proof
  showcases the special treatment required to model many of the most common
  features of dependent type theories, thus making our proof applicable to a
  broad range of type systems.
\item We strengthen our logical relation to prove the existence of
  $\beta$-normal forms (Section~\ref{sec:extension}) and $\beta\eta$-normal
  forms (Section~\ref{sec:betaeta}) for well-typed open terms. The
  modifications made to our initial logical relation are small and closely
  mirror the necessary extensions for a simply-typed language. We use this
  part to show that once we have established the base techniques, we can port
  ideas from proofs about simpler languages to the dependently typed setting.
\item We mechanize all our proofs using the Coq proof assistant, with 957
  lines of code for the consistency proof and a moderate increase to 1568
  lines of code for the normalization proof.  We discuss the specifics related
  to our choice of Coq as our metatheory, including our use of off-the-shelf
  semantic engineering infrastructure and automation tools, in
  Section~\ref{sec:logrelmech}.  Our proof scripts, with comments, are
  available to reviewers as supplementary materials.
\item We compare our work to existing proofs by logical relations and other
  proof techniques for proving consistency and normalization.  We provide an
  overview of this prior work (Section~\ref{sec:relatedwork}) and also give an
  in-depth explanation on how various design decisions affect the size of our
  proof and its extensibility to additional features
  (Section~\ref{sec:discuss}).
\end{itemize}

@@

The syntax of \lang can be found in Figure~\ref{fig:syntax}.  As a dependent
type theory, terms and types are collapsed into the same syntactic
category. The type $[[Set i]]$ represent a universe type, annotated by its
universe level, a natural number $i$.  Abstractions $[[ \ x . a ]]$ and
dependent function types $[[Pi x : A . B]]$ are binding forms for the variable
$x$ in the body of the function and codomain of the function type.
\footnote{In the exposition in this paper, binding forms are equal up to
  $\alpha$-conversion and we adopt the Barendregt Variable
  Convention~\cite{barendregt:lambda-calculus}, which lets us assume that
  bound variables are distinct.  In some places, we are informal about the
  treatment of variables and substitution; our mechanized proofs make these
  notions precise by using de Bruijn indices~\cite{debruijn1994automath}. }
We use the notation $[[A -> B]]$ when the output type $[[B]]$ is not dependent
on the input variable. For simplicity, we omit the type annotations in the abstraction
forms. We discuss how the inclusion of type annotations can affect
our development in Section~\ref{sec:betaeta}, where we extend our
consistency result to the existence of $\beta\eta$-normal forms.
We include in \lang{} the intensional identity type
$[[a ~ b : A]]$ whose proofs can be eliminated by the J-eliminator
$[[J c a b p]]$, where $[[p]]$ is an equality proof between $[[a]]$ and
$[[b]]$, and $[[c]]$ is the term whose type is to be casted.  Finally, \lang{}
includes booleans, with standard syntax.  \scw{Example of what you can use
  identity types for? Or an example of a program that uses J?}
